Supplementary Materials

### 2022-12-27 manuscript/supplementary.tex
ms notes

\begin{align*}
  n_T &= \frac{n(n-1)}{2} \\
  n_S &= {c_1 \choose 2} + {c_2 \choose 2} \\
  n_D &= n_T - n_S
\end{align*}
%
which would be stored in row $\mathbf{Y}_l = [n_S, n_D]$.

\subsection{Pairwise Diversity Estimates}

The pairwise diversity at site $l$ across the $n$ sampled chromosomes can be
calculated from the $\mathbf{Y}$ matrix as follows,

\begin{align}
  \pi(l) = \frac{n_D}{n_T}
\end{align}

@@

\begin{align}
  \pi(l) = \frac{2}{n(n-1)} \sum_{i < j}^n k_{i,j}
\end{align}

where $k_{i,j}$ is 1 if the alleles at this site differ, and 0 otherwise. Note
that for sites with missing data, $c_1 = c_2 = 0$ and thus $n_T = 0$ and the
division is invalid. Such cases were marked as missing with floating point
values \texttt{NaN}s. Binned summaries of data are weighed by the number of
complete cases, e.g. rows with $c_1 + c_2 > 0$.

This per-basepair diversity is then averaged over all callable bases, to give a
genome-wide or window estimate of diversity of accessible bases in the set
$\mathcal{A}$,

\begin{align*}
  \bar{\pi}(\mathcal{A}) = \frac{\sum_{l \in \mathcal{A}} \pi(l)}{L_\mathcal{A}}.
\end{align*}

where $L_\mathcal{A} = |\mathcal{A}|$ is the number of accessible bases. Note
that the sum in the numerator is random over the sample of chromosomes sampled
from the population. Estimates of pairwise diversity often condition on the
accessible bases, and thus treat this as fixed. However, the number of
accessible bases varies across the chromosome; this can lead to a source of
apparent bias during block-bootstrap estimates of uncertainty.  In this case,
pairwise diversity is a ratio estimator, and is thus biased, since by Jensen's
inequality $\E(\nicefrac{y}{x}) \ge \nicefrac{\E(y)}{\E(x)}$ for random
variables $x$ and $y$.

The genome-wide $\bar{\pi}$ can also be estimated by summing the columns of
$\mathbf{Y}$, and calculating $\nicefrac{n_D}{n_T}$.

@@

\section{}

>>>>

\begin{align*}
  n_T &= \frac{n(n-1)}{2} \\
  n_S &= {c_1 \choose 2} + {c_2 \choose 2} \\
  n_D &= n_T - n_S
\end{align*}
%
which would be stored in row $\mathbf{Y}_l = [n_S, n_D]$. Note that the
per-site $\mathbf{Y}$ handles non-polymorphic sites and missing data.
Non-polymorphic sites have $n_S = {n \choose 2}$ and $n_D = 0$, and missing
data has $n_S = n_D = 0$.

\subsection{Pairwise Diversity Estimates}
\label{supsec:diversity}

The pairwise diversity at site $l$ across the $n$ sampled chromosomes can be
calculated from row $l$ of the $\mathbf{Y}$ matrix as follows,

\begin{align}
  \pi_l = \frac{n_D}{n_T}
\end{align}

@@

\begin{align}
  \pi_l = \frac{2}{n(n-1)} \sum_{i < j}^n k_{i,j}
\end{align}

where $k_{i,j}$ is 1 if the alleles at this site differ at site $l$, and 0
otherwise.

There are three ways to aggregate $\pi_l$ across all sites. The first is,

\begin{align}
  \label{eq:}
  \pi^{(1)} = \frac{1}{L} \sum_{i=1}^L \frac{n_{D,i}}{n_{T,i}}
\end{align}

which if the number of samples across loci is constant, simplifies to an
unweighted average across sites. Second, one can take a weighted average, with
weights determined by the total number of samples present at a site,

\begin{align}
  \label{eq:}
  \pi^{(2)} = \frac{1}{\sum_{i=1}^L n_i} \sum_{i=1}^L n_i \frac{n_{D,i}}{n_{T,i}}
\end{align}

Third, one can weight by the number of pairwise comparisons at a site,
$n_{T,i}$, rather than total number of samples, $n_i$, which leads to a ratio
of sums,

\begin{align}
  \label{eq:}
  \pi^{(3)} = \frac{\sum_{i=1}^L n_{D,i}}{\sum_{i=1}^L n_{T,i}}.
\end{align}

We predominantly use the estimator $\pi^{(3)}$, as it corresponds to how we
summarize the matrix $\mathbf{Y}$ across windows for our likelihood. All
methods have mean squared errors and biases very close to one another (TODO).

Note, however, that estimates of pairwise diversity often condition on the
accessible bases, and thus treat this as fixed. However, the number of
accessible bases varies across the chromosome; this can lead to a source of
apparent bias during block-bootstrap estimates of uncertainty.  In this case,
pairwise diversity is a ratio estimator, and is thus biased, since by Jensen's
inequality $\E(\nicefrac{y}{x}) \ge \nicefrac{\E(y)}{\E(x)}$ for random
variables $x$ and $y$.

@@

\section*{Likelihood}

Our model is essentially a generalized nonlinear model with a binomial link
function. This is the form used by \textcite{Elyashiv2016-vt} and
\textcite{Murphy2022-sj}. These models fit the observed number of pairwise
nucleotide differences in genomic windows to the expected pairwise diversity
under some evolutionary model. We only consider BGS models here, so the mean
function for position $x$ is the product of the proportion by which BGS reduces
diversity at position $x$, $B(x)$, and genome-wide neutral diversity $\pi_0$,

\begin{align}
  \pi(x) = B(x, \Phi) \pi_0
\end{align}

$\Phi$ is the set of BGS parameters (i.e. the DFE for each feature type and
mutation rate), and $\pi_0 = 4 N_e \mu$ is determined by the genome-wide
drift-effective population size $N_e$, set by only reproductive and demographic
processes. Regional mutation rate heterogeneity can be accounted for with a
regional mutation rate scaling function $m(x)$, $\pi(x) = B(x, \Phi) m(x)
\pi_0$, but we do not explore mutation-rate heterogeneity, as it is unclear how
to differentiate variance in mutation rate from the substitution rate
heterogeneity we find across the genome.

The likelihood of the set of parameters $\Psi = \{\pi_0, \Phi\}$ can be written
in the form of,

\begin{align}
  \log\mathcal{L}(\Psi) = \sum_{v \in \mathcal{V}} \sum_{i \ne j \in \mathcal{S}} \log(P(O_{i,j}(v) | \Psi))
\end{align}

(c.f. \cite{McVicker2009-ax,Elyashiv2016-vt,Murphy2022-sj}) where $\mathcal{V}$
is the set of putatively neutral sites, $\mathcal{S}$ is the set of samples,
and $\Psi$ are the BGS parameters. The indicator variable $O_{i,j}(v)$ is 1 if
samples $i$ and $j$ are different at putatively neutral site $v$, and zero
otherwise. While the theoretic $\pi(v)$ gives the average number of pairwise
differences, for small values, this is approximately the heterozygosity
probability, so we can write

\begin{align}
  P(O_{i,j}(v) | \Psi) = 
    \begin{cases}
      \pi(v), & O_{i,j}(v, \Psi) = 1 \\
      1-\pi(v), & O_{i,j}(v, \Psi) = 0 \\
    \end{cases}.
\end{align}

(c.f. \cite{Elyashiv2016-vt}).

As in Section \ref{supsec:diversity}, the number of pairwise differences and
the total number of pairwise comparison at a site are sufficient statistics for
the likelihood. Then, the binomial log-likelihood for the data at site $v$ is,

\begin{align}
  \ell_v(\Psi) &= \log(\pi(v, \Psi)) n_{\text{D},v} + \log(1-\pi(v, \Psi)) n_{\text{S},{v}}
\end{align}

\subsection{The scale of processes}

We can observe $\widehat{\pi}(x)$ at a per-basepair resolution. However, for a
variety of reasons, we do not want to fit the composite likelihood model to the
per-basepair scale of data. First, this would be computationally infeasible.
Second, the mean function $\pi(x)$ varies on a natural scale that is itself a
free parameter of the model. Our model can be written as,

\begin{align}
  \ell(\Psi, h) &= \sum_{b} \sum_{v \in \mathcal{V}_b} \ell_v(\Psi) \\
             &= \sum_{b} \left[\log(\bar{\pi}(b, \Psi)) \sum_{v \in \mathcal{V}_b} n_{\text{D},v} + \log(1-\bar{\pi}(b, \Psi)) \sum_{v \in \mathcal{V}_b} n_{\text{S},{v}}\right] \\
             &= \sum_{b} \left[\log(\bar{\pi}(b, \Psi)) Y_{\text{D},b} + \log(1-\bar{\pi}(b, \Psi)) Y_{\text{S},{b}}\right] \\
\end{align}

where $h$ is the bandwidth or window size, $b$ are physical-scale window
indices, $\bar{\pi}(b | \Psi)$ are the average diversity in bin $b$, and
$Y_{\text{S},b}$ and $Y_{\text{S},b}$ are the sums across putatively neutral
sites within a bin. We fit $h$ using out-sample cross validation.

This corresponds to a binomial likelihood for the observed data summarized at
genomic scale $h$. Thus an alternative way to express this model is as,

\begin{align}
  Y_{\text{D},b} \sim \text{Binom}(\bar{\pi}(b, \Psi), Y_{\text{D},b} + Y_{\text{S},b}).
\end{align}

Here, $\bar{\pi}(b, \Psi)$ is assumed to be the \emph{probability} of sampling
two different alleles, rather than the average \emph{number} of pairwise
differences; these are approximately equal when $\pi$ is small. This
corresponds to an identity link function; one could alternatively use a
two-alleles finite sites model link function of the form, $\pi/(1 + 2 \pi)$. We
experimented with this and found there was little difference between these link
functions, so we opted for the simpler identity link function.

\subsection{The Background Selection Reduction Factor}

Under the background selection model, the diversity in window $b$ is $\pi(b,
\Psi) = \bar{B}(b, \Psi) \pi_0$. Here, $\bar{B}(b | \Psi)$ is the predicted
reduction in diversity due to BGS in window $b$, given background selection
parameters $\Psi$. In practice, we pre-calculate $\bar{B}(b | \Psi)$ at fixed
sites across the genome, and take the average of the fixed sites $B$ values
within a widow.

In practice, this is site-specific. We can write the
reduction at any neutral site $v$ in the genome as the product of $B$s across
all segments,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \int f(\mu(\mathcal{A}(g)), s, S_g) w(s|\mathcal{A}(g)) ds \right)
\end{align}

where $S_g$ is exogenous genomic data about the segment, $S_g = \{L_g, r_g,
\rho(|v-p_g|)\}$, where $L_g$ is the segment's length, $r_g$ is the
recombination rate per basepair in the segment, and $\rho(|v-p_g|)$ is the
recombination distance between the focal site $v$ and the segment position
$p_g$ (we approximate, and use the nearest end position to the neutral site).
Additionally, $w(s|\mathcal{A}(g))$ is the distribution of selection
coefficients for segment $g$, if segment $g$ is a member of annotation class
$\mathcal{A}(g)$.

We can think about the DFE as the conditional distribution of a particular
selection coefficient given a mutation occurs, for a particular annotation
class. The BGS function $f(\cdot)$ only depends on $\mu$ through the
introduction of deleterious alleles with selection coefficient $s$ at rate
$\omega(s|\mathcal{A}(g)) = \mu(\mathcal{A}(g)) w(s | \mathcal{A}(g))$. Thus,
we can write,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \int f(\omega(s|\mathcal{A}(g)), s, S_g) ds \right)
\end{align}

which we can discretize as,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \sum_s f(\omega(s|\mathcal{A}(g)), s, S_g) \right).
\end{align}

Next, note that there are a finite number of annotation classes, $\mathcal{A}
\to \{a_1, a_2, \ldots, a_k\}$, so we can further partition this as

\begin{align}
  B(v | \theta) = \exp\left(- \sum_{\{g \;:\; \mathcal{A}(g) = a_1\}} \sum_s f(\omega(s| a_1), s, S_g) + \sum_{\{g \;:\; \mathcal{A}(g) = a_2\}} \sum_s f(\omega(s| a_2), s, S_g) + \ldots  \right)
\end{align}

Let us define the $d_\omega \times d_s \times d_g$ multidimensional array
$\mathbf{F}$, and the $d_g \times d_a$ feature classification matrix
$\mathbf{A}$.

\section{Calculation of B Maps}

Each B map is determined by the annotation track of putatively conserved
segments, which includes (1) the segment positions and lengths, (2) the type of
feature of each segment (e.g.  exons, UTRs, phastcons conserved elements,
etc.), as well as (3) the recombination map, and (3) the deleterious selection
coefficient $s(i)$ and mutation rate at which these deleterious enter the
population for feature type $i$.

The B maps are computed at fixed, evenly spaced sites.

\section{Ratchet Rate Prediction}

\begin{align}
  R = \int 
\end{align}
